---
title: "2015 Projections"
author: "Neeraj Asthana"
date: "May 21, 2016"
output: pdf_document
---

#2015 Fantasy Football Projections

##Summary

This notebook is looking to predict player performance in the 2015 Fantasy Football season based on how players are projected to perform. 

##Reading dataset

The data set was obtained from:

I wrote a scraping script in python and then used the find all feature in Sublime to remove all of the quotations.

Basic cleaning to get numerics in the columns

```{r setup, include=FALSE}
setwd("/home/neeraj/Documents/Projects/FantasyFootball/data/formatted_data")
data <- read.csv("2015espnprojstats.csv", header = TRUE, na.strings = "--")
data[,1] <- unlist(lapply(data[,1], as.character))

for (i in 5:16){
  data[,i] <- as.numeric(data[,i])
}

#what fantasy players actually scored and performed
actual <- read.csv("2015stats.csv")

#dataset containing ages and all pro selections
ages <- read.csv("2015ages.csv", stringsAsFactors=FALSE)
```

###Size of the dataset

```{r}
print(dim(data))
```

###Column names

```{r}
head(data)

head()
```

##Visualizations

